---
title: "Longitudinal Final Project"
author: "Meg Rosales"
date: "4/19/2020"
output: word_document
params:
  t: !r Sys.time()
---
```{r}
#print current date and time
params$t

#working directory
setwd("C:/Users/meg/Desktop/Courses/Spring 2020 Courses/Longitudinal Data Analysis")

#packages
library(readxl)
library(dplyr)
library(tidyr)
library(lme4)
library(ggplot2)
library(lmerTest)
library(lattice)
library(naniar)

```

**Data Cleaning**
I saved the cleaning code so you could see my process, but I generated the data files from this and then saved them so I wouldn't have to rerun this every time since the cleaning required some manual entry.
```{r echo = F}

# #don't need to run this because the files are saved at the end of this chunk and I just load them
# #read in the data
# phon <- read_excel("KS SPSS data copy.xlsx use for combofinal.xlsx")
# 
# #rename variables - add 9 to tccount9 and tcorplos9 and 39 to concount39 and tccount39
# names(phon) <- edit(names(phon))
# 
# #variables for analysis
# #concount: number of consonants in child's consonant inventory
# #tccount: number of true consonants in child's consonant inventory
# #tcorplos: percentage of correct stop consonants produced
# selnames <- c("subject","cleftype","concount9","tccount9","tcorplos9","concount13","tccount13","tcorplos13",
#  "concount17","tccount17","tcorplos17","concount21","tccount21","tcorplos21","concount27","tccount27", 
# "tcorplos27","concount33","tccount33","tcorplos33","concount39","tccount39","tcorplos39")
# 
# #narrow down to variables for analysis
# cons <- phon%>%
#   select(selnames)
# 
# #recode cleftype to binary cleft variable: 1 = cleft lip/palate, 0 = no cleft
# #5 = no cleft, 1-4 = cleft
# cons$cleft <- ifelse(cons$cleftype==5, 0, 1)
# 
# #need data for three different models
# #consonant inventory data
# cinv <- cons%>%
#   select(c("subject", "cleft", "concount9", "concount13", "concount17", "concount21", "concount27", "concount33",
#            "concount39"))
# 
# #true consonant inventory data
# tc <- cons%>%
#   select(c("subject", "cleft", "tccount9", "tccount13", "tccount17", "tccount21", "tccount27", "tccount33",
#            "tccount39"))
# 
# #total correct plosives data
# #only look at from 17 to 39 months because babies don't make enough consonants/correct consonants to analyze
# #this until about 17 months
# tcp <- cons%>%
#   select(c("subject", "cleft","tcorplos17","tcorplos21","tcorplos27","tcorplos33", "tcorplos39"))
# 
# #drop anyone missing a cleft status
# #dataset had data mostly complete for one site as some of the data was from a different site
# #in the interest of time, I only used the cases from the completed site
# cinv <- cinv%>% drop_na(cleft)
# tc <- tc%>% drop_na(cleft)
# tcp <- tcp%>% drop_na(cleft)
# 
# #Particip 100 has no data, drop them
# cinv <- cinv%>% filter(subject != 100)
# tc <- tc%>% filter(subject != 100)
# tcp <- tcp%>% filter(subject != 100)
# 
# #the rest of the data seems to be missing completely at random <- no relationship to group
# 
# #save the datasets so we don't have to clean them again
# cinv <- save(cinv, file = "cinv.Rdata")
# tc <- save(tc, file = "tc.Rdata")
# tcp <- save(tcp, file = "tcp.Rdata")
```

**Data Analysis**

1) Consonant inventory analysis: Does size of overall consonant inventory at each age differ between children with a cleft and their non-cleft peers? Is there a difference in trends over time for size of overall consonant inventory between children with a cleft and their non-cleft peers?

***Approach:*** Linear mixed effects model
          ****Outcome:****  concount = (continuous) Size of child's consonant inventory
          ****Fixed effect:**** cleft = (binary) Whether child has a cleft (1) or not (0)
          ****Random effect:**** subject
          
```{r}
#read in the data
load("cinv.Rdata")

#plot of means over time

#get the mean counts by group into a dataframe
cinv_m <- cinv%>%
  group_by(cleft)%>%
  summarise(mean9 = round(mean(concount9, na.rm = T)),
            mean13 = round(mean(concount13, na.rm = T)),
            mean17 = round(mean(concount17, na.rm = T)),
            mean21 = round(mean(concount21, na.rm = T)),
            mean27 = round(mean(concount27, na.rm = T)),
            mean33 = round(mean(concount33, na.rm = T)),
            mean39 = round(mean(concount39, na.rm = T)))

means_cinv <- reshape(as.data.frame(cinv_m),
                 varying = c("mean9", "mean13", "mean17", "mean21", "mean27", "mean33","mean39"),
                 v.names = "mean",
                 timevar = "months",
                 times = c(9, 13, 17, 21, 27, 33, 39),
                 direction = "long")%>%
  mutate(cleft = as.factor(cleft))


#create time plot with mean lead level vs. time
ggplot(data = means_cinv, aes(x = months, y = mean, group = cleft, color = cleft))+
  geom_line()+
  xlab("Age in months")+
  ylab("Mean size of consonant inventory")+
  geom_point(shape = 1, size = 4)
```

```{r}
#restructure data for mixed effects model

#change from tibble to dataframe because otherwise reshape complains
cinv <- as.data.frame(cinv)

#mixed effects model restructuring
cinv_me <- reshape(cinv,
                   varying = c("concount9", "concount13", "concount17", "concount21", "concount27", "concount33","concount39"),
                   v.names = "concount",
                   timevar = "months",
                   times = c(9, 13, 17, 21, 27, 33, 39),
                   direction = "long")%>%
  #get rid of superfluous id variable this generates
  select(-c(id))%>%
  #convert months to a factor variable - only 6 time points
  mutate(months = as.factor(months))

#linear mixed effects model
cinv_model_lm <- lmer(concount ~ cleft*months + (1 | subject), 
             data = cinv_me)
summary(cinv_model_lm)
```

2) True consonant inventory analysis: Does the size of the true consonant inventory at each age differ between children with a cleft and their non-cleft peers? Is there a difference in trends over time for the size of the true consonant inventory between children with a cleft and their non-cleft peers?

****Approach:**** Linear mixed effects model
          ****Outcome:****  tccount = (continuous) Number of true consonants in child's consonant inventory 
          ****Fixed effect:**** cleft = (binary) whether child has a cleft (1) or not (0)
          ****Random effect:**** subject
          
```{r}
#read in the data
load("tc.Rdata")
#get the mean counts by group into a dataframe
tc_m <- tc%>%
  group_by(cleft)%>%
  summarise(mean9 = round(mean(tccount9, na.rm = T)),
            mean13 = round(mean(tccount13, na.rm = T)),
            mean17 = round(mean(tccount17, na.rm = T)),
            mean21 = round(mean(tccount21, na.rm = T)),
            mean27 = round(mean(tccount27, na.rm = T)),
            mean33 = round(mean(tccount33, na.rm = T)),
            mean39 = round(mean(tccount39, na.rm = T)))

means_tc <- reshape(as.data.frame(tc_m),
                 varying = c("mean9", "mean13", "mean17", "mean21", "mean27", "mean33","mean39"),
                 v.names = "mean",
                 timevar = "months",
                 times = c(9, 13, 17, 21, 27, 33, 39),
                 direction = "long")%>%
  mutate(cleft = as.factor(cleft))


#create time plot with mean lead level vs. time
ggplot(data = means_tc, aes(x = months, y = mean, group = cleft, color = cleft))+
  geom_line()+
  xlab("Age in months")+
  ylab("Mean size of true consonant inventory")+
  geom_point(shape = 1, size = 4)

```
```{r}
#restructure data for mixed effects model

#change from tibble to dataframe because otherwise reshape complains
tc <- as.data.frame(tc)

#mixed effects model restructuring
tc_me <- reshape(tc,
                   varying = c("tccount9", "tccount13", "tccount17", "tccount21", "tccount27", "tccount33","tccount39"),
                   v.names = "tccount",
                   timevar = "months",
                   times = c(9, 13, 17, 21, 27, 33, 39),
                   direction = "long")%>%
  #get rid of superfluous id variable this generates
  select(-c(id))%>%
  #convert months to a factor variable - only 6 time points
  mutate(months = as.factor(months))

#linear mixed effects model
tc_model_lm <- lmer(tccount ~ cleft*months + (1 | subject), 
             data = tc_me)
summary(tc_model_lm)
```

3) Percent total correct plosives analysis: Does percentage correct of stop consonants produced at each age differ between children with a cleft and their non-cleft peers? Is there a difference in trends over time for percentage correct of stop consonants produced between children with a cleft and their non-cleft peers?

**Approach:** Linear mixed effects model
          **Outcome:**  concount = (continuous) Number of consonants in child's consonant inventory at age
          **Fixed effect:** cleft = (binary) whether child has a cleft (1) or not (0)
          **Random effect:** subject (want to make inference generalizable outside this population, interpret at                                the subject level)
          
```{r}
#read in the data
load("tcp.Rdata")

#need to recode percentages of 101 to NA (a convention they had when entering the data)
tcp <- tcp%>%replace_with_na_all(condition = ~.x == 101.00)

#get the mean counts by group into a dataframe
tcp_m <- tcp%>%
  group_by(cleft)%>%
  summarise(mean17 = mean(tcorplos17, na.rm = T),
            mean21 = mean(tcorplos21, na.rm = T),
            mean27 = mean(tcorplos27, na.rm = T),
            mean33 = mean(tcorplos33, na.rm = T),
            mean39 = mean(tcorplos39, na.rm = T))

means_tcp <- reshape(as.data.frame(tcp_m),
                 varying = c("mean17", "mean21", "mean27", "mean33","mean39"),
                 v.names = "mean",
                 timevar = "months",
                 times = c(17, 21, 27, 33, 39),
                 direction = "long")%>%
  mutate(cleft = as.factor(cleft))


#create time plot with mean lead level vs. time
ggplot(data = means_tcp, aes(x = months, y = mean, group = cleft, color = cleft))+
  geom_line()+
  xlab("Age in months")+
  ylab("Mean percent correct stop consonants produced")+
  geom_point(shape = 1, size = 4)

```

```{r}
#restructure data for mixed effects model

#change from tibble to dataframe because otherwise reshape complains
tcp <- as.data.frame(tcp)

#mixed effects model restructuring
tcp_me <- reshape(tcp,
                   varying = c("tcorplos17", "tcorplos21", "tcorplos27", "tcorplos33","tcorplos39"),
                   v.names = "tcorplos",
                   timevar = "months",
                   times = c(17, 21, 27, 33, 39),
                   direction = "long")%>%
  #get rid of superfluous id variable this generates
  select(-c(id))%>%
  #convert months to a factor variable - only 6 time points
  mutate(months = as.factor(months))

#linear mixed effects model
tcp_model_lm <- lmer(tcorplos ~ cleft*months + (1 | subject), 
             data = tcp_me)
summary(tcp_model_lm)
```